---
title: "Chapter 19: Agent-based Modeling -- Empirically Calibrated"
published-title: r_recap
engine: knitr
freeze: auto
bibliography: literature.bib
csl: ASA.csl
---

## Introduction

*Note: this script has been kindly provided by [Sascha Grehl](https://www.uni-leipzig.de/en/profile/mitarbeiter/sascha-grehl). You might notice the difference in coding style -- i.e., old pipe operator, not necessarily using all of the `tidyverse` packages. Style preferences differ, but this script impressively showcases that there are different ways and they are neither mutually exclusive nor necessarily better. I made some changes when I deemed them necessary, mostly to ensure compatibility with Macs. You can find the original code [in this GitHub repository](https://github.com/SGrehl/diffusion/tree/main).*

Welcome to our session on Agent-Based Modeling (ABM) and parameter calibration in R. In the coming chapters, you will learn how to set up a simple ABM that simulates the diffusion of a superior innovation among agents, then use calibration techniques to estimate model parameters from observed (or synthetic) data.

Agent-Based Modeling (ABM) is a computational approach where a large number of individual entities -- the agents -- interact within a defined environment. Each agent can follow simple or complex rules, and through these interactions, emergent phenomena arise at the system level. The agents can represent people, firms, animals, or any discrete decision-making units. Rules are pre-specified by the model architect and often include how agents move, communicate, adopt behaviors, or otherwise change over time. Finally, the environment can be spatial (e.g., a grid or network), social (e.g., hierarchy, peer groups), or both.

ABMs are nice for a couple of reasons: they capture heterogeneity as ach agent can have different attributes or decision rules. Furthermore, they model interactions directly -- agents can influence each other’s state (e.g., spread of ideas or contagion). Finally, they allow for emergence, thereby explicitly capturing the final micro--macro step of the Coleman boat; this allows us to thoroughly model and investigate how large-scale patterns (e.g., diffusion curves) arise from simple local interactions.

However, a model, no matter how cleverly designed, is most useful when it can be compared to real data or synthetic data that mimics reality. Calibration is the process of adjusting a model’s internal parameters so that its output matches observed data as closely as possible. To this end, we perform parameter calibration, i.e., find the combination of parameters (e.g., contagion probability, contact rate, etc.) that leads to the most realistic model outputs. Then we can can validate: once the parameters are estimated, we need to check if the model can predict data in scenarios beyond what it was originally fit to. This is useful since a calibrated model is more reliable for exploring what-if scenarios or interventions.

In today's session, we will use synthetic historic data generated with hidden (“secret”) parameters. Your goal will be to figure out which parameters best explain the simulated outcomes—mimicking what you would do with real-world data if these parameters were unknown. To this end, we will first build up an ABM. Then, we will build a function that takes two models and compares the goodness of their fit (i.e., how well their outcomes match). Finally, we will systematically alter the parameters of our existing model to find the specification that matches our existing, synthetic data the best. 

By the end of this session, you will have practical experience in:

- Building and running an agent-based simulation in R.
- Implementing calibration routines and interpreting the results.
- Visualizing parameter search results with heatmaps or other techniques.

## Building the model

In this chapter, we will delve into the mechanics of an agent-based model (ABM) that simulates the diffusion of an innovation over time. Our model starts with a grid-based world of agents, allows them to interact and spread the innovation, and then collects data on how the innovation diffuses over time.

For our diffusion simulation, we use a simple mesh grid of size $M \times M$. Each cell on this grid is occupied by $n$ agents. Agents are either innovators (already using the new innovation) or non-innovators.

At each time step, all innovators:
1. Draw a random distance `d` from a specified distribution (e.g., geometric, Weibull, etc.) using the `model_implement_distance_function`  
2. Select one random cell exactly `d` units away (in Manhattan distance).  
3. Pick one random agent in that cell (the targeted agent).  
4. Attempt to convince the targeted agent to adopt the innovation, succeeding with probability `p` (defined by `agent_distance_contagion`).  

If successful, the targeted agent becomes an innovator as well.

Additionally, we include a `seed` for reproducibility, so each run can produce consistent results. Here is an example `list` capturing these model parameters:

```{r}
needs(tidyverse)
model <- list(
  seed                     = 12345,       # for reproducibility
  world_size               = 13,          # grid dimension (M)
  agents_per_cell    = 6,           # agents per cell (n)
  agent_distance_function  = "geometric", # the specified distribution d is drawn from
  agent_distance_parameter = c(0.8),      # the parameters for the distribution function
  agent_adoption_p         = 0.8          # adoption probability (p)
)
```

### Creating the World

After defining the model parameters, we need a function to initialize our world with agents. Each agent is placed on a \( M \times M \) grid, with \( n \) agents per cell. We also specify which agent starts off with the innovation.

```{r}
world_init <- function(world_size = 9,
                       agents_per_cell = 3) {
  # Create the grid
  cells <- expand.grid(x = 1:world_size, y = 1:world_size)
  
  # Assign agents to grid cells
  world <- cells %>%
    slice(rep(1:n(), each = agents_per_cell)) %>%
    mutate(
      id       = row_number(),
      cell_id  = x + (y - 1) * world_size,
      # Initialize the innovation to 1 in the center for just one agent
      innovation = ifelse(
        x == ceiling(world_size / 2) & 
        y == ceiling(world_size / 2) & 
        row_number() == ceiling((world_size * world_size * agents_per_cell) / 2),
        1, 
        0
      )
    )
  return(world)
}

world <- world_init()
world |> head()
```

Now the world is a data frame, for each agent a row with its `x` and `y` coordinates, the `cell_id`, and its `innovation` status (`1 == innovator`, `0 == not`).

### Additional Setup: Distance Drawing and Position Lookup

Before we will turn to updating the World in each step, we will introduce to functions which will help us to run the simulation faster.  

When we run our ABM, each innovator needs to pick a distance `d` from a probability distribution (e.g., a geometric or Weibull distribution). To avoid repeatedly switching among these distributions at each step, we define a single helper function that attaches the appropriate `draw_distance()` function to the `model` object.

```{r}
model_implement_distance_function <- function(model){
  model$draw_distance <- switch(
    model$agent_distance_function,
    "weibull" = function(n = 1) {
      round(rweibull(n,
                     shape = model$agent_distance_parameter[1],
                     scale = model$agent_distance_parameter[2]))
    },
    "geometric" = function(n = 1) {
      round(rgeom(n, prob = model$agent_distance_parameter[1]))
    },
    stop("Unsupported agent distance function")
  )
  return(model)
}

model_implement_distance_function(model)
```

We also define a position lookup function to quickly find all agents in a specific cell, greatly improving performance when we need to identify potential targets.

```{r}
create_position_lookup <- function(world) {
  # Group by cell_id, store a list of all IDs in that cell
  # 'deframe()' makes a named list with 'cell_id' as the name
  world %>%
    group_by(cell_id) %>%
    summarize(agent_ids = list(id), .groups = "drop") %>%
    deframe()
}

create_position_lookup(world) |> head()
```

### A single time step

At each time step in our ABM, we want to simulate how existing innovators attempt to convince other agents to adopt the innovation. The process includes:

1. Identifying all current innovators.
2. Drawing a random distance for each innovator and selecting a target cell using Manhattan distance.
3. Picking a random agent in that cell.
4. Updating the targeted agents' status if they have been convinced to adopt the innovation.

However, for performance reasons, our actual implementation filters out would-be unsuccessful innovators before drawing distances. This approach avoids unnecessary computations for innovators who would fail anyway. For example, if we have 1,000 innovators but only a 5% chance of convincing someone, filtering first means we only process about 50 innovators for the subsequent steps -- significantly reducing computational overhead.

Below is the `world_step()` function, highlighting how these steps are organized for both correctness and efficiency.

```{r}
world_step <- function(
  world, 
  model, 
  position_lookup = create_position_lookup(world) # if no position_lookup is provided, it is calculated automatically
){
  
  # 1) Select all agents who are innovators
  innovators <- world %>% filter(innovation == 1)
  
  # If everyone is already converted, don't run the simulation step
  if (nrow(innovators) == nrow(world)) return(world)
  
  # 2) Filter out innovators whose random_value is above the threshold. 
  #    That is, only keep those who will successfully convince a target agent.
  #    Filtering here saves performance because we skip the following steps 
  #    for innovators who wouldn't convert anyone anyway.
  innovators <- innovators %>%
    mutate(random_value = runif(n())) %>%
    filter(random_value > model$agent_adoption_p)
    
  # If no innovators remain after filtering, skip the step
  if (nrow(innovators) == 0) return(world)
  
  # 3) Draw a random distance for each innovator
  innovators <- innovators %>%
    mutate(
      distances = model$draw_distance(nrow(innovators))
    )
  
  # 4) For each innovator, determine a random target cell using Manhattan distance
  innovators <- innovators %>%
    mutate(
      target_cell = mapply(
        function(x, y, d) {
          get_a_cell_at_manhattan_distance(x, y, d, model$world_size)
        },
        x, y, distances
      )
    )
  
  # 5) For each innovator, pick a random target agent within the target cell
  innovators <- innovators %>%
    mutate(
      target_id = mapply(
        function(tc, my_id) {
          # If invalid cell, return -1
          if (tc == -1) return(-1)
          
          # Possible targets in that cell
          targets_id <- position_lookup[[tc]]
          
          # Exclude the innovator itself
          targets_id <- targets_id[targets_id != my_id]
          
          # If no one is left in that cell, return -1
          if (length(targets_id) == 0) return(-1)
          
          # Otherwise pick one at random
          sample(targets_id, 1)
        },
        target_cell,  # from innovators
        id            # from innovators
      )
    )
  
  # Collect all valid targets (target_id != -1)
  targets <- innovators %>% filter(target_id != -1) %>% pull(target_id)
  
  # 6) Update those target agents to become innovators
  if (length(targets) > 0) {
    world$innovation[world$id %in% targets] <- 1
  }

  return(world)
}

world_step(world, model) |> 
    head()
```

One key detail is how we sample a random cell at distance `d`. We use the Manhattan metric (often referred to as L1 distance), which measures distance by summing the absolute differences in the `x` and `y` coordinates. In a grid setting, moving two cells to the left is considered equivalent in “distance cost” to moving one cell left and one cell up -— both yield a Manhattan distance of 2. This makes sense for grid-based worlds, where movement is often discrete rather than diagonal or Euclidean.

The function for this is as follows:
```{r}
get_a_cell_at_manhattan_distance <- function(x,
                                             y,
                                             d, 
                                             world_size
) {
  # Return the same cell right away if d == 0
  if (d == 0) return(x+(y-1)*world_size)
  
  # For these values, we know for sure, there are no valid cells
  if (d > (world_size-1)*2) return(-1)
  
  cells_at_distance <- do.call(
    rbind,
    lapply(0:d, function(k) {
      dist_rem <- d - k
      # Four possible points for each k
      rbind(
        c(x + k, y + dist_rem),
        c(x + k, y - dist_rem),
        c(x - k, y + dist_rem),
        c(x - k, y - dist_rem)
      )
    })
  )
  
  # Remove duplicates (especially when k=0 or dist_rem=0)
  cells_at_distance <- unique(cells_at_distance)
  
  # Filter out-of-bounds points
  in_bounds <- cells_at_distance[,1] >= 1 & cells_at_distance[,1] <= world_size &
    cells_at_distance[,2] >= 1 & cells_at_distance[,2] <= world_size
  cells_at_distance <- cells_at_distance[in_bounds, , drop = FALSE]
  
  # If no valid cells remain return -1
  if (nrow(cells_at_distance) == 0) return(-1) 
  
  # Pick one cell at random
  chosen_cell <- cells_at_distance[sample(nrow(cells_at_distance), 1), ]
  
  return(chosen_cell[1]+(chosen_cell[2]-1)*world_size)
}

```

### Running the Full Simulation

After defining how to initialize the world (`world_init()`) and how a single time step unfolds (`world_step()`), we now need to combine these pieces to run our ABM across multiple rounds. The function `run_simulation()` orchestrates all the steps, producing a list of world states -— one for each time step.

Here is the function:
```{r}
run_simulation <- function(model, 
                           max_steps = 25
){
  set.seed(model$seed)
  worlds <- list()

  for (i in 1:max_steps){
    if (i==1) {
      # 1) Initialize the world in the first iteration
      worlds[[i]] <- world_init(model$world_size, model$agents_per_cell)
      
      # 2) Create a lookup table for quick agent lookups
      #    Agents do not move, so building this once is sufficient.
      position_lookup <- create_position_lookup(worlds[[i]]) 
      
      # 3) Attach the chosen distance-drawing function to the model
      #    (e.g., "weibull" or "geometric")
      model <- model_implement_distance_function(model)
    } else {
      # 4) Update the world one step at a time
      worlds[[i]] <- world_step(worlds[[i-1]], model, position_lookup)
    }
  }
  return(worlds)
}
```

### Plotting the Simulation

Understanding the dynamics of your agent-based model (ABM) is crucial. Visualizations help you interpret and communicate the spread of innovation across the grid. In this section, we'll explore two primary methods for visualizing the simulation results: Static plots provide snapshot views of the world at specific time steps. Animations, on the other hand, give dynamic visualizations the show the evolution of the innovation over time.

#### Creating Static Plots

A static plot provides a clear snapshot of the world's state at a particular time step. The `world_print()` function uses `ggplot2` to generate a color-coded grid that represents the share of innovators in each cell.
    
```{r}
world_print <- function(world, 
                        step = NA,
                        palette = "RdYlBu") {
  # Summarize innovation per cell
  world_summary <- world %>%
    group_by(x, y) %>%
    summarise(share_innovation = mean(innovation), .groups = 'drop')
  
  # Create the plot
  plot <- ggplot(world_summary, aes(x = x, y = y, fill = share_innovation)) +
    # Draw a black background for the entire grid
    geom_tile(data = expand.grid(x = seq(min(world$x), max(world$x)), y = seq(min(world$y), max(world$y))), aes(x = x, y = y), fill = "black") +
    # Overlay the tiles for non-empty cells
    geom_tile() +
    scale_x_continuous(breaks = 1:max(world_summary$x)) +
    scale_y_continuous(breaks = 1:max(world_summary$y)) +
    coord_fixed() +
    theme_minimal() +
    scale_fill_distiller(palette = palette, limits = c(0, 1)) +  # Fill color for tiles
    labs(title = paste0("World Innovation Map: Round ", step),
         x = "X Coordinate",
         y = "Y Coordinate")
  
  print(plot)
}

sim_data <- run_simulation(model)

world_print(sim_data[[3]], step = 3)
world_print(sim_data[[25]], step = 25)
```

#### Creating Animations

While static plots are useful for snapshots, animations provide a comprehensive view of how innovation spreads over time. Animations can make it easier to identify patterns, trends, and anomalies that might not be apparent from static images.

Due to server restrictions on the University Cluster's RStudio Server version, certain animation functionalities may not work directly within the environment. To circumvent this, animation functions are excluded from the main script and placed in a separate file: `functions_animations.R`. This allows you to run animations locally on your machine without interfering with server policies. 

*(Note Felix: the following function has not been tested on the server. You might consider using Sascha's original function that can be found in the [GitHub repository](https://github.com/SGrehl/diffusion/tree/main).)*

```{r, eval=TRUE}
needs(gganimate)

world_animation <- function(worlds){
    all_worlds <- bind_rows(lapply(seq_along(worlds), function(i) {
    worlds[[i]] %>%
      group_by(x, y) %>%
      summarise(share_innovation = mean(innovation), .groups = 'drop') %>%
      mutate(time = i)
  }))

    anim <- ggplot(all_worlds, aes(x = x, y = y, fill = share_innovation)) +
    geom_tile() +
    scale_x_continuous(breaks = 1:max(all_worlds$x)) +
    scale_y_continuous(breaks = 1:max(all_worlds$y)) +
    coord_fixed() +
    theme_minimal() +
    scale_fill_distiller(palette = "RdYlBu", limits = c(0, 1)) +
    labs(title = "World Innovation Map: Step {frame_time}") +
    transition_time(time) +
    ease_aes('linear')
  
    animate(anim, nframes = length(worlds), fps = 2)

}

world_animation(sim_data)
```

## Calibration

In this chapter, we delve into calibration, a crucial process in agent-based modeling (ABM) that aligns simulation outputs with observed (or synthetic) data. Calibration ensures that your model accurately reflects the real-world processes it aims to represent by fine-tuning its parameters. We will explore how to measure the closeness of your simulation to historic data and employ techniques to identify the best-fitting parameters.

While our simulation generates a wealth of detailed data, this richness can be overwhelming and, in many cases, unnecessary for calibration purposes. To effectively calibrate our model, we first need to create a summary of the simulation data to a form that is comparable to our observed (historic) data. This involves extracting the relevant metrics from the simulation that correspond to the data available from the real-world process we aim to model. This ensures that we are comparing like with like, making the calibration process meaningful and effective. We will refer to this data reduction process as "generating a report."" The purpose of the report is to distill the simulation's rich data into a concise summary that mirrors the structure and content of our real-world data.

In our script, the `report_generate(world_history)` function takes a list of world states at different time steps (referred to as `world_history`) and produces a data frame containing all relevant information necessary for calibration.

#### Example Report Structure

For our diffusion of innovation model, we assume that we know:

- The global share of innovators in the population.
- The number of cells in the grid that fall into specific share thresholds of innovators.

Here's an illustrative example of what the report might look like:

```{r}
report_generate <- function(world_history){
  report <- tibble()
  
  for (i in seq_along(world_history)) {
    
    world <- world_history[[i]]
    
    share_global <- sum(world$innovation, na.rm = TRUE) / nrow(world)
    
    world_summary <- world %>%
      group_by(x, y) %>%
      summarise(share_innovation = mean(innovation), .groups = 'drop') %>%
      mutate(share_innovation = round(share_innovation, 3) |> as.character()) |> 
      group_by(share_innovation) %>%
      summarise(cells = n(), .groups = 'drop') %>%
      mutate(step = i) %>% 
      pivot_wider(
        names_from = share_innovation, 
        values_from = cells,
        values_fill = 0  # Fill missing values with 0
      ) %>% 
      ungroup() %>% 
      mutate(share_global = share_global)
    
    report <- bind_rows(report, world_summary)
    
  }
  
  report <- report %>%
    mutate(across(everything(), ~ replace_na(., 0))) %>% 
    relocate(step, share_global)
  
  share_cols <- names(report)[!names(report) %in% c("share_global", "step")]  # Exclude original columns
  new_names <- paste0("s_", seq_along(share_cols))              # Generate s_1, s_2, ...
  names(report)[!names(report) %in% c("share_global", "step")] <- new_names
  
  return(report)
}

report_generate(sim_data) |> head()
```

The report is structured as follows: 

- `step` represents the time step in the simulation.
- share_global: The overall proportion of innovators in the entire population.
- `s_1` to `s_7`: The number of cells that fall into specific thresholds of local innovation shares: `s_1` are the cells with 0% innovators; `s_2` to `s_7`: Cells with 1/6 up to 6/6 (100%) innovators.

For instance, at step 1, the global share of innovators is 0.000986, with 168 cells having 0% innovators and 1 cell with 1/6 innovators. By step 6, the global share increases to 0.00690, and the distribution of innovators across cells becomes more varied.

Including this summarized information is essential for calibration because it mirrors the structure of the real-world data we aim to fit our model to. In the next section, we will discuss how to quantify the fit between the simulation reports and historic data using a fit function.

### Properties of a good fit function

A fit function (also known as an objective function, distance metric, or error measure) quantifies how closely the simulation’s output aligns with the observed (historic) data. Selecting an appropriate fit function is critical for effective calibration. Here are the desirable properties of a good fit function:

- Dimensionless:
  - The fit function should be dimensionless to allow the comparison of discrepancies across different data dimensions.
  - This enables the aggregation of errors without bias towards any particular scale.
- Weighted:
  - The ability to assign weights to different data points ensures that more important aspects of the data have a greater influence on the fit score.
  - This flexibility allows prioritizing certain metrics over others based on their relevance.
- Concave:
  - The fit function should treat two small discrepancies of size \( a \) as more desirable than one large discrepancy of size \( 2a \) and no discrepancy elsewhere.
  - This encourages a more uniform fit across all data points rather than allowing a single outlier to dominate the fit score.
- Symmetric:
  - The function should treat positive and negative discrepancies equally.
  - For example, being twice as high as the observed value should carry the same weight as being half as high.
- Non-Negative:
  - Discrepancies should be non-negative to ensure that errors do not cancel each other out.
  - This property maintains the integrity of the fit score as a true measure of discrepancy.
- Finite:
  - The fit function should produce finite values for finite inputs.
  - This ensures stability and reliability in the calibration process.

A commonly used fit function that satisfies these properties is:

$$
Fit = \sum_{i=1}^v w_i \cdot \left( \frac{h_i−m_i}{\frac{h_i+m_i}{2}} \right) ^2
$$

With

-  $h_i$ = Historic (observed) value $i$.
-  $m_i$ = Model (simulation) value $i$.
-  $w_i$ = Weight assigned to this particular data point $i$. 
-  $v$   = Number of data points fitted.

This function computes the squared relative difference, ensuring symmetry, non-negativity, and dimensionlessness. The weight w allows for the prioritization of specific data points. However, to achieve this, the minimum values of h and m must be set to 0, which may require some data transformation. Then a perfect fits is always a Fit of 0.

### Defining the Fit Function

To quantitatively assess the similarity between our simulation output and historic data, we define the `measure_fit()` function. This function calculates the fit score by comparing the relevant metrics extracted from both datasets.

```{r}
# Example measure of fit between two reports
measure_fit <- function(historic_report, sim_report) {
  # Join by step (inner join: only compare matching steps)
  df <- dplyr::inner_join(historic_report, sim_report,
                          by = "step",
                          suffix = c("_hist", "_sim"))
  
  # Identify numeric columns to compare (excluding "step")
  compare_cols <- setdiff(names(df), "step")
  compare_cols_hist <- grep("_hist$", compare_cols, value = TRUE)
  compare_cols_sim  <- grep("_sim$",  compare_cols, value = TRUE)
  
  # We'll assume the columns match in order, i.e. 
  # for every X_hist there is a corresponding X_sim
  # For safety, you could match them by sub-string if needed.
  
  # For each pair, compute squared differences
  diff <- c()
  for (col_hist in compare_cols_hist) {
    col_sim <- sub("_hist$", "_sim", col_hist)  # get the matching sim column
    if (col_sim %in% names(df)) {
      # numeric difference
      v_hist <- df[[col_hist]]
      v_sim  <- df[[col_sim]]
      # you can use a different method to calculate the difference
      # but this function has the following property:
      # if the min.value is 0, regardless of the max.value the diff is in the interval [0,4], this makes it very comparable
      diff <- c(diff, ((v_hist - v_sim)/((v_hist +v_sim)/2))^2) 
      diff[is.nan(diff)] <- 0
    }
  }
  
  # Return mean difference (or sum, or RMSE, etc.)
  mean(diff, na.rm = TRUE)
}

```

As described earlier, we might want to give different weight to certain parameters; this function allows for this. We can either give more weight to particular steps or particular columns.
```{r}
# a more complex function that allows for different weights with respect to certain cols or steps
measure_fit_weighted <- function(historic_report, 
                                 sim_report, 
                                 weight_steps = NULL,    # e.g. c(1,2,3,4,5)
                                 weight_cols  = NULL    # e.g. c("share_global"=2, "s_1"=1, "s_2"=1, ...)
) {
  # Join by step (inner join: only compare matching steps)
  df <- dplyr::inner_join(
    historic_report, sim_report,
    by = "step", suffix = c("_hist", "_sim")
  )
  
  # Identify numeric columns to compare (excluding "step")
  compare_cols <- setdiff(names(df), "step")
  compare_cols_hist <- grep("_hist$", compare_cols, value = TRUE)
  compare_cols_sim  <- grep("_sim$",  compare_cols, value = TRUE)
  
  # We will store our weighted differences in a vector:
  all_diff <- numeric(0)  # empty numeric vector
  
  for (col_hist in compare_cols_hist) {
    # Find the matching simulated column
    col_sim <- sub("_hist$", "_sim", col_hist)
    if (!(col_sim %in% names(df))) next  # skip if missing
    
    # Extract the underlying variable name (without _hist / _sim)
    var_name <- sub("_hist$", "", col_hist)
    
    # Grab vectors from the data frame
    v_hist <- df[[col_hist]]
    v_sim  <- df[[col_sim]]
    
    # We'll compute the unweighted difference
    raw_diff <- ((v_hist - v_sim)/((v_hist + v_sim)/2))^2
    
    # Replace NaN with 0
    raw_diff[is.nan(raw_diff)] <- 0
    
    # ------------------------------------------------------------------
    # Column-based weighting
    # ------------------------------------------------------------------
    # If `weight_cols` is provided, it should be something like:
    #   weight_cols = c("share_global"=2, "s_1"=1, "s_2"=3, ...)
    # If var_name isn't in weight_cols, we'll assume weight = 1
    if (!is.null(weight_cols) && var_name %in% names(weight_cols)) {
      w_col <- weight_cols[[var_name]]
    } else {
      w_col <- 1
    }
    
    # ------------------------------------------------------------------
    # Step-based weighting
    # ------------------------------------------------------------------
    # If `weight_steps` is provided, e.g. weight_steps = c(1,2,3,4,5)
    # We assume the step values go from 1..N, so we can do:
    #   w_step[i] = weight_steps[df$step[i]]
    # If no match, default to 1.
    
    # Create a vector of step weights, same length as raw_diff
    if (!is.null(weight_steps)) {
      # Some caution: step might not start at 1. 
      # If your steps are strictly 1..max, we can do:
      w_steps <- weight_steps[df$step]
      # If step is not guaranteed to be an integer index, you can do 
      # a named lookup, e.g. weight_steps = c("1"=1, "2"=2, ...)
      # w_steps <- weight_steps[as.character(df$step)]
      # fallback for missing names
      w_steps[is.na(w_steps)] <- 1
    } else {
      w_steps <- rep(1, length(raw_diff))
    }
    
    # ------------------------------------------------------------------
    # Combine the weights: multiply raw_diff by (w_col * w_step)
    # ------------------------------------------------------------------
    weighted_diff <- raw_diff * w_col * w_steps
    
    # Accumulate in our all_diff vector
    all_diff <- c(all_diff, weighted_diff)
  }
  
  # Finally, compute your desired summary (mean, sum, RMSE, etc.)
  mean(all_diff, na.rm = TRUE)
}
```

The calibrate_model function tests different parameter combinations in an agent-based model to find settings that best match historical data. It creates a grid of parameter combinations, including grid sizes, agents per cell, distance functions, and adoption probabilities. For each combination, it runs multiple simulations to account for randomness, generates reports comparing to historic data, and calculates an average fit score.

The function works by first building a data frame of all parameter combinations using `expand.grid()`. It then loops through each combination, modifying the base model with new parameters using `modifyList()`. For each parameter set, it runs the simulation `n_sims` times with different random seeds, generates reports using `report_generate()`, and measures the fit against historic data using `measure_fit()`. The results are stored in a tibble showing how well each parameter combination performed.

```{r}
needs(foreach, doParallel)

calibrate_model <- function(
    historic_report,
    model,
    # if model variable is FALSE, use default provided by `model`, otherwise use them to generate a grid
    world_size               = NA,
    world_agents_per_cell    = NA,
    agent_distance_function  = NA,
    agent_distance_parameter = NA,    
    agent_adoption_p  = NA,
    # number of simulations runs for each combination
    n_sims = 3,
    # number of steps in each run
    max_steps = 25
) {
  # 1) Build a data frame of all combinations
  param_grid <- expand.grid( 
    world_size               = if_else(is.na(world_size)              , model$world_size              , world_size),
    world_agents_per_cell    = if_else(is.na(world_agents_per_cell)   , model$world_agents_per_cell   , world_agents_per_cell),
    agent_distance_function  = if_else(is.na(agent_distance_function) , model$agent_distance_function , agent_distance_function),
    agent_distance_parameter = if_else(is.na(agent_distance_parameter), model$agent_distance_parameter, agent_distance_parameter),
    agent_adoption_p         = if_else(is.na(agent_adoption_p)        , model$agent_adoption_p        , agent_adoption_p),
    stringsAsFactors = FALSE
  )
  
  # We'll store results in a data frame
  results <- tibble()
  
  # 2) Loop over all combinations in the grid
  for (i in seq_len(nrow(param_grid))) {
    tmp_model <- modifyList(model, as.list(param_grid[i, ]))
    
    print(param_grid[i, ])

    fit_values <- numeric(n_sims)  # holds error score for each replicate
    
    # 3) Repeat the simulation multiple times to average out randomness
    for (rep_i in seq_len(n_sims)) {
      # Increase seed to vary each run
      tmp_model$seed <- model$seed - 1 + rep_i
      
      # Run the simulation
      sim_worlds <- run_simulation(tmp_model, max_steps)
      
      # Generate the simulated report
      sim_report <- report_generate(sim_worlds)
      
      # Compute error/fit relative to historic data
      fit_values[rep_i] <- measure_fit(historic_report, sim_report)
    }
    
    # Average fit score across replicates
    avg_fit <- mean(fit_values)
    
    # 4) Store the results
    results <- results %>%
      bind_rows(tibble(
        world_size               = tmp_model$world_size,
        world_agents_per_cell    = tmp_model$world_agents_per_cell,
        agent_distance_function  = tmp_model$agent_distance_function,
        agent_distance_parameter = tmp_model$agent_distance_parameter,
        agent_adoption_p         = tmp_model$agent_adoption_p,
        start_seed               = model$seed,
        simulations              = n_sims,
        fit = avg_fit
      ))
  }
  
  return(results)
}

```

Since this is tedious work that might take a long time, this parallel version of the calibration function, `calibrate_model_mc()`, uses multiple cores to speed up parameter testing. It creates a parallel computing cluster, distributes work across available cores, and combines results into a single dataset. The implementation improves efficiency when testing many parameter combinations.

```{r}
calibrate_model_mc <- function(
    historic_report, 
    model,
    # if model variable is FALSE, use default provided by `model`, otherwise use them to generate a grid
    world_size               = NA,
    world_agents_per_cell    = NA,
    agent_distance_function  = NA,
    agent_distance_parameter = NA,    
    agent_adoption_p         = NA,
    # number of simulations runs for each combination
    n_sims = 3,
    # number of steps in each run
    max_steps = 25,
    # maximum cores to use 
    max_cores = 24
) {
  # 1) Build a data frame of all combinations
  param_grid <- expand.grid( 
    world_size               = if_else(is.na(world_size)              , model$world_size              , world_size),
    world_agents_per_cell    = if_else(is.na(world_agents_per_cell)   , model$agents_per_cell   , world_agents_per_cell),
    agent_distance_function  = if_else(is.na(agent_distance_function) , model$agent_distance_function , agent_distance_function),
    agent_distance_parameter = if_else(is.na(agent_distance_parameter), model$agent_distance_parameter, agent_distance_parameter),
    agent_adoption_p         = if_else(is.na(agent_adoption_p)        , model$agent_adoption_p        , agent_adoption_p),
    stringsAsFactors = FALSE
  )
  
  # 2) Set up parallel back-end
  n_cores <- min(parallel::detectCores(), max_cores, nrow(param_grid))
  cat("Using", n_cores, "cores\n")
  
  # Initialize a cluster and register it
  cl <- parallel::makeCluster(n_cores)
  doParallel::registerDoParallel(cl)
  
  # 3) Use foreach to iterate in parallel over each row of param_grid
  #    .combine = rbind => combine results row-by-row
  #    .packages = "dplyr" => loads dplyr on each worker
  #    .export => exports functions from global environment
  #               that the workers need (run_simulation, report_generate, measure_fit)
  
  results <- foreach(
     i = seq_len(nrow(param_grid)),
    .combine  = rbind,
    .packages = c("dplyr"),
    .export   = c("run_simulation", "report_generate", "measure_fit")
  ) %dopar% {
    source("files/functions_simulation.R")
    
    # Copy the base model and override parameters from row i
    tmp_model <- modifyList(model, as.list(param_grid[i, ]))
    
    # Run multiple simulations for this combination
    fit_values <- numeric(n_sims)
    for (rep_i in seq_len(n_sims)) {
      # Vary the seed slightly for each replicate
      tmp_model$seed <- model$seed - 1 + rep_i
      
      sim_worlds  <- run_simulation(tmp_model, max_steps)
      sim_report  <- report_generate(sim_worlds)
      fit_values[rep_i] <- measure_fit(historic_report, sim_report)
    }
    
    # Average fit score across replicates
    avg_fit <- mean(fit_values)
    
    # Return a one-row data.frame for foreach to combine
    data.frame(
      world_size               = tmp_model$world_size,
      world_agents_per_cell    = tmp_model$world_agents_per_cell,
      agent_distance_function  = tmp_model$agent_distance_function,
      agent_distance_parameter = tmp_model$agent_distance_parameter,
      agent_adoption_p         = tmp_model$agent_adoption_p,
      start_seed               = model$seed,
      simulations              = n_sims,
      fit                      = avg_fit,
      stringsAsFactors         = FALSE
    )
  }
  
  # 4) Stop the cluster
  parallel::stopCluster(cl)
  
  # Convert results to a tibble and return
  results <- as_tibble(results)
  return(results)
}
```

The `calibration_heatmap` function creates a ggplot2 visualization showing how model fit varies across parameter combinations. It displays results as a color-coded tile plot where the x and y axes represent different parameters and the color intensity indicates fit quality. The function offers customization of variables, color schemes, and plot styling.
```{r}
calibration_heatmap <- function(
    data,
    x_var       = "agent_distance_parameter",   # Name of the column to use for x-axis
    y_var       = "agent_adoption_p",   # Name of the column to use for y-axis
    fill_var    = "fit",                       # Name of the column to use for the fill (e.g. fit/error)
    palette     = "RdYlBu",                    # Brewer palette name (see ?RColorBrewer)
    direction   = -1,                          # 1 or -1, flips the palette direction
    plot_title  = "Calibration Heatmap"
) {
  ggplot(data = data, aes_string(x = x_var, y = y_var, fill = fill_var)) +
    geom_tile() +
    scale_fill_distiller(palette = palette, direction = direction) +
    labs(
      title = plot_title,
      x     = x_var,
      y     = y_var,
      fill  = fill_var
    ) +
    theme_minimal()
}
```

So now let's have a look at our historic data. The next chunk loads our data and animates its innovation process. This serves as our "ground truth."

```{r}
load("files/historic_data.RData")

historic_report <- report_generate(historic_data)
world_animation(historic_data)
```

In the next step, we first compare the historic data with our prior run. For this, we measure the fit at each time-step. Perfect overall fit -- i.e., if our model perfectly replicates the patterns in the historic data, would be 0.
```{r eval=TRUE}
source("files/functions_simulation.R")
source("files/functions_calibration.R")

load("files/historic_data.RData")

# run a single model

model_sim <- list(
  seed                     = 1234,
  world_size               = 13,
  world_agents_per_cell    = 6,
  agent_distance_function  = "geometric",   
  agent_distance_parameter = 0.5,
  agent_adoption_p         = 0.5
)

sim_data <- run_simulation(model_sim)
sim_report <- report_generate(sim_data)

measure_fit(historic_report, sim_report)
```

Then we can move on to running multiple (i.e., 50 per parameter combination) models whose outcomes we subsequently compare to the "real," observed outcome. For this comparison, we draw a heatmap. 
```{r eval=FALSE} 

# define the base model
base_model <- list(
  seed                     = 1234,
  world_size               = 13,
  world_agents_per_cell    = 6,
  agent_distance_function  = "geometric",   
  agent_distance_parameter = NA_real_,    # placeholder
  agent_adoption_p         = NA_real_     # placeholder
)

# run the calibration
calibration_result <- calibrate_model_mc(
  historic_report          = historic_report,
  model                    = base_model,
  agent_distance_parameter = seq(0.05, 1.0, by = 0.05),
  agent_adoption_p         = seq(0.05, 1.0, by = 0.05),
  n_sims                   = 50,
  max_steps                = 25
)

save(calibration_result, file =  "files/calibration_result.Rdata")

# generate a heatmap based on calibration
calibration_heatmap(calibration_result)
```


```{r}
#| warning: false
#| echo: false

load("files/calibration_result.Rdata")
calibration_heatmap(calibration_result)
```

The bluer the tile, the better the fit of our own model to the "original" data. According to our graph, the true adoption parameters lie around 0.4, and the true distance parameter around 0.6.

While in this toy example, the "ground truth" came from and unspecified simulation run of our own, in real research applications you would want to compare your simulation outcomes to data that stem from observational, longitudinal data (e.g., surveys, observational studies).

## Further links

- Another handbook entry on [ABMs in the social sciences](https://doi.org/10.4337/9781789909432.00011)
- A [workshop on ABMs](https://marijnkeijzer.github.io/sub/bigsss-abm/) by Marijn Keizer and Jan Lorenz (it's in Netlogo but gives you some infos on how to build and interpret ABMs)