---
title: "Chapter 8: `OpenAI whisper`"
published-title: whisper
engine: knitr
freeze: false
bibliography: literature.bib
csl: ASA.csl
---

In this chapter, I will show you how to use `OpenAI Whisper` for audio transcription and diarization. Whisper is a versatile tool that helps convert audio recordings into text, lending itself well for tasks like transcribing interviews, radio shows, or any other type of recorded speech. Additionally, we will use speaker diarization to identify different speakers in the audio.

Throughout this chapter, we will use `reticulate` for integrating Python code within our R workflow, `pydub` for audio manipulation, `openai-whisper` for audio transcription, `torch` for running deep learning models, `numpy` for numerical operations, and `pyannote.audio` for speaker diarization.

## Install Python using `reticulate` and `miniconda`

```{r echo=FALSE, message=FALSE, warning=FALSE}
vembedr::embed_youtube("KKhGNHgLJm8")
```

Like in the chapter on `selenium`, we first create a `conda` environment with all the required packages.

```{r eval=FALSE}
needs(reticulate)
#reticulate::conda_create(envname = "pyenv/whisper_env") # create empty environment
reticulate::conda_install(envname = "pyenv/whisper_env",
                          packages = c("pydub", "openai-whisper", "torch", "numpy"), 
                          pip = TRUE) # install packages into a certain environment
```

Thereafter, we need to make sure that we activate our environment. Moreover, for `pyannote`, you will need an access token from huggingface (get yourself a "read" key here: https://huggingface.co/settings/tokens). I stored mine in my R environment and feed it forward to Python, so that it is readily accessible without others seeing it. 

```{r eval=FALSE}
needs(reticulate)
use_condaenv(condaenv = "pyenv/whisper_env")
```

Then we can load the required Python packages.

```{python eval=FALSE, include=FALSE}
hf_token = "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"
```

```{python eval=FALSE}
import torch
import whisper
from pyannote.audio import Pipeline
import wave
import os
from pydub import AudioSegment
from pyannote.core import Segment
import numpy as np
import pandas as pd
from scipy.io import wavfile
```

Also, if we have a GPU available -- as you probably should if you're running these operations on a server -- we need to tell our packages that they can use the GPU instead of the CPU. 

```{python eval=FALSE}
device = torch.device("mps" if torch.backends.mps.is_available() else "cpu") # I'm using MPS here because I'm on a Mac
# device = device = torch.device('cuda' if torch.cuda.is_available() else 'cpu') # if you're on a server with a suitable GPU

whisper_model = whisper.load_model("base").to("cpu")
#whisper_model = whisper.load_model("base").to(device) # if you're on a server with a suitable GPU

# Load diarization pipeline
pipeline = Pipeline.from_pretrained("pyannote/speaker-diarization", use_auth_token = hf_token)
pipeline.to(device)
```

Since this script is for students of Leipzig University first and foremost, I had to make some changes to it so that it can run nicely on the university's server. It's quite a chore to install `ffmpeg` there -- which `whisper` requires by default to read in sound. Here, we skip this step, but due to this the transcription function needs to be rewritten from scratch. However, this comes with the shortcoming of the model in this script only accepting .wav files. This can be circumnavigated by using the `audioread` library (find more infos [here](https://github.com/beetbox/audioread)).   

```{python eval=FALSE}
def load_audio_manually(file_path, target_sr=16000):
    sr, audio = wavfile.read(file_path)
    if audio.ndim > 1:
        audio = np.mean(audio, axis=1)
    if sr != target_sr:
        import librosa
        audio = librosa.resample(audio.astype(np.float32), orig_sr=sr, target_sr=target_sr)
    return audio

def transcribe_audio(file_path): #transcribe it
    audio = load_audio_manually(file_path)
    audio = audio.astype(np.float32) / 32768.0  # Normalize if original was int16
    result = whisper_model.transcribe(audio)
    return result
```

Now we canmake our first transcription:

```{python eval=FALSE}
ihaveadream_transcript = transcribe_audio("mlk_ihaveadream.wav")
ihaveadream_transcript["text"]
ihaveadream_transcript["segments"][0]
```

## Diarization

```{r echo=FALSE, message=FALSE, warning=FALSE}
vembedr::embed_youtube("seSdYnCR4Z0")
```

We can also use speaker diarization to split audio files by speaker and transcribe each segment separately. This is particularly useful if the recording contains multiple speakers. Here, we use the `AudioSegment` class from `pydub` to load the audio file. The `pipeline` object is used to iterate over the speaker turns, and we save each speaker segment to a separate audio file.

```{python eval=FALSE}
audio = AudioSegment.from_wav("temp/thisisamerica_200_snippet.wav")

for turn, _, speaker in pipeline("temp/thisisamerica_200_snippet.wav").itertracks(yield_label=True):
    start_time = turn.start * 1000
    end_time = turn.end * 1000
    segment_audio = audio[start_time:end_time]
    segment_file = f"temp/carrboro_market/{speaker}-{int(turn.start)}-{int(turn.end)}.wav"
    segment_audio.export(segment_file, format = "wav")
    #print(f"Saved segment: {segment_file}")
```

Next, we want to transcribe each of the diarized segments and save the results to a CSV file.

```{python eval=FALSE}
import pandas as pd
import glob

# Initialize lists to store each attribute separately
speakers = []
start_times = []
end_times = []
texts = []

def transcribe_and_collect(file_path, speaker, start_time, end_time):
    # Perform transcription
    result = transcribe_audio(file_path)  # Assuming transcribe_audio function exists
    # Append each attribute to its respective list
    speakers.append(speaker)
    start_times.append(start_time)
    end_times.append(end_time)
    texts.append(result['text'])


# Iterate over diarized segments (assuming you have diarization data)
for segment_file in glob.glob("temp/carrboro_market/SPEAKER_*.wav"):
    parts = segment_file.split('-')
    speaker = parts[0].split("/")[2]
    start_time = float(parts[1])
    end_time = float(parts[2].split('.')[0])
    transcribe_and_collect(segment_file, speaker, start_time, end_time)

# Write the DataFrame to a CSV file
transcriptions_df = pd.DataFrame({
    "speaker": speakers,
    "start": start_times,
    "end": end_times,
    "text": texts
})

transcriptions_df
# Write the DataFrame to a CSV file
transcriptions_df.to_csv("temp/transcriptions.csv", index=False)
```

So that we can finally read it back in and wrangle the results in R.

```{r}
transcriptions_df <- readr::read_csv("temp/transcriptions.csv") |> 
  dplyr::arrange(start)

transcriptions_df
```

So here we are, a quick introduction to transcription and diarization in Python.

## Further links

-   `whisper` [GitHub repository](https://github.com/openai/whisper)
-   `Pyannote` [audio documentation](https://pyannote.github.io/)

